Retrofitting Effect Handlers onto OCaml

### 2021-03-24 retro-concurrency.tex
Minor proofread edits

\begin{enumerate}[label=R\arabic*]
	\item \textbf{Backwards compatibility.} Extant OCaml programs \kc{Extant
		better than legacy?} do not break under OCaml extended with effect
		handlers. OCaml code that does not use effect handlers will pay negligible
		performance and memory cost.

@@

The function |spawn| (line 9) evaluates the computation |f| in an effect
handler. The computation |f| may return normally with a value, or perform
effects |Fork f'| and |Yield|. The pattern |effect Yield k| handles the effect
|Yield| and binds |k| to the continuation of the corresponding |perform|
delimited by this handler. The scheduler queue |runq| maintains a queue of
these continuations. |suspend| pushes continuations into the queue, |run_next|
pops continuations from the queue and resumes them with |()| value using the
|continue| primitive. In the case of |Yield| effect, we suspend the current
continuation |k| and resume the next available continuation. In the case of
|Fork f'| effect, we suspend the current continuation and recursively call
|spawn| on |f'| in order to run |f'| concurrently. Observe that we can change
the scheduling algorithm from FIFO to LIFO by changing the scheduler queue to a
stack.

@@

The interaction of non-local control flow with systems programming is quite
subtle~\cite{TFP17, Leijen18}. Consider the following function that uses
blocking I/O functions from the OCaml standard library to copy data from the
input channel |ic| to the output channel |oc|:
\begin{lstlisting}
let copy ic oc =
  let rec loop () =
	  output_string oc ((input_line ic) ^ "\n");
		loop ()
	in
	try loop () with
  | End_of_file -> close_in ic; close_out oc
	| e -> close_in ic; close_out oc; raise e
\end{lstlisting}
The function |input_line| raises an |End_of_file| exception on reaching the end
of input, which is handled by the exception handler which closes the channels.
The |close_*| functions do nothing if the channel is already closed. The code
is written in a defensive style to handle other exceptional cases such as the
channels being closed externally. Both |input_line| and |output_string| raise
|Sys_error| exception if the channel is closed. In this case, the catch-all
exception handler closes the channels and reraises the exception to communicate
the exceptional behaviour to the caller.

>>>>

\begin{enumerate}[label=R\arabic*]
	\item \textbf{Backwards compatibility.} Extant OCaml programs \kc{Extant
		better than legacy?} \tk{Existing might be better?} do not break under
    OCaml extended with effect handlers. OCaml code that does not use effect
    handlers will pay negligible performance and memory cost.

@@

The function |spawn| (line 9) evaluates the computation |f| in an effect
handler. The computation |f| may return normally with a value, or perform
effects |Fork f'| and |Yield|. The pattern |effect Yield k| handles the effect
|Yield| and binds |k| to the continuation of the corresponding |perform|
delimited by this handler. The scheduler queue |runq| maintains a queue of
these continuations. |suspend| pushes continuations into the queue, |run_next|
pops continuations from the queue and resumes them with |()| value using the
|continue| primitive. In the case of the |Yield| effect, we suspend the current
continuation |k| and resume the next available continuation. In the case of the
|Fork f'| effect, we suspend the current continuation and recursively call
|spawn| on |f'| in order to run |f'| concurrently. Observe that we can change
the scheduling algorithm from FIFO to LIFO by changing the scheduler queue to a
stack.

@@

The interaction of non-local control flow with systems programming is quite
subtle~\cite{TFP17, Leijen18}. Consider the following function that uses
blocking I/O functions from the OCaml standard library to copy data from the
input channel |ic| to the output channel |oc|:
\begin{lstlisting}
let copy ic oc =
  let rec loop () =
	  output_string oc ((input_line ic) ^ "\n");
		loop ()
	in
	try loop () with
  | End_of_file -> close_in ic; close_out oc
	| e -> close_in ic; close_out oc; raise e
\end{lstlisting}
The function |input_line| raises an |End_of_file| exception on reaching the end
of input, which is handled by the exception handler which closes the channels.
The |close_*| functions do nothing if the channel is already closed. The code
is written in a defensive style to handle other exceptional cases such as the
channels being closed externally. Both |input_line| and |output_string| raise
a |Sys_error| exception if the channel is closed. In this case, the catch-all
exception handler closes the channels and reraises the exception to communicate
the exceptional behaviour to the caller.
